Are mental properties supervenient on brain properties?

### 2011-08-05 nature_sci_rep/submission2/supervenience.tex
updated refs
finished redline cep comments of aug 5 2011

\noindent Questions and assumptions about mind-brain supervenience go back at least as far as Plato's dialogues in circa 400 BCE \cite{Plato97}.  While there are many different notions of supervenience, we find Davidson's canonical description particularly illustrative \cite{Davidson70}:
%  The mind-brain supervenience notion was canonized in 1970 with the following quote from Donald Davidson \cite{Davidson70}:
\begin{quotation}
\noindent [mind-brain] supervenience might be taken to mean that there cannot be two events alike in all physical respects but differing in some mental respect, or that an object cannot alter in some mental respect without altering in some physical respect.
\end{quotation}
This philosophical conjecture has potentially widespread implications.  
% We consider a special case of supervenience of considerable interest.  Specifically, this work addresses a version of a local supervenience between mental properties and brain properties, with particular emphasize on brain-graphs (i.e., connectivity structure).  Supervenience questions or assumptions are fundamental to 
% The determination of supervenience (or lack thereof) of minds on brains has potentially important implications in 
% a number of fields of inquiry.  
For example, neural network theory and artificial intelligence often implicitly assume 
% take a generalized notion of 
a local version mind-brain supervenience 
% as an assumption 
% , which if falsified, might change modern approaches to learning 
\cite{Haykin2008,Ripley2008}. Cognitive neuroscience similarly seems to operate under such assumptions
% , which if falsified, might result in novel perspectives and theories 
\cite{Fodor1998, Gazzaniga2008}.  Philosophers continue to debate and refine notions of supervenience 
% And the question of mind-brain supervenience continues to be debated amongst philosophers 
\cite{Kim2007}.  
% Moreover, many cognitive neuroscience investigations could be cast as questions
Yet, to date, relatively scant attention has been paid to what might be empirically learned about supervenience.

@@

% Ideally, 
Based on the above considerations,
we might consider the following hypothesis test: $H_0: L^*>0$ and $H_A: L^*=0$; rejecting the null indicates that $\MsB$. Unfortunately, 
% because the null hypothesis is not closed, 
the alternative hypothesis lies on the boundary, so the $p$-value is always equal to unity \cite{Bickel2000}.  From this, Theorem \ref{thm:2} follows immediately:
\begin{thm} \label{thm:2}
	There does not exist a viable test of $\MsB$
\end{thm}

@@

\subsection*{The existence of a consistent statistical test for $\eps$-supervenience} % (fold)

@@

Unfortunately, the rate of convergence of $L_{F}(g_n)$ to $L_{F}(g^*)$ depends on the (unknown) distribution $F=F_{MB}$ \cite{DGL96}. Furthermore, arbitrarily slow convergence theorems regarding the rate of convergence of $L_{F}(g_n)$ to $L_{F}(g^*)$ demonstrate that there is no universal $n,n'$ which will guarantee that the test has power greater than any specified target $\beta > \alpha$ \cite{Devroye83}. For this reason, the test outlined above can provide only a one-sided conclusion: if we reject we can be $100(1-\alpha)$\% confident that $\MeB$ holds, but we can never be confident in its negation; rather, it may be the case that the evidence in favor of $\MeB$ is insufficient 
% for any number of reasons, including that 
because we simply have not yet collected enough data. 
% Thus, without restrictions on $F_{MB}$, arbitrarily slow convergence theorems imply that our theorem of $\varepsilon$-supervenience does not \tr{strictly} satisfy Popper's {\it falsifiability} requirement \cite{Popper}.

\subsection*{The availability of a consistent statistical test for $\eps$-supervenience} % (fold)
\label{sub:uc}

The above considerations indicate the existence of a consistent test for $\eps$-supervenience whenever the classifier used is consistent.  
% that to construct a statistical test with guaranteed limiting unity power, a universally consistent classifier (one for which $\hL^{n'}_F(g_n) \conv L_F(g^*)$ as $n,n' \conv \infty$) is required. 
To actually implement such a test, one must be able to (i) measure mind/brain pairs and (ii) have a consistent classifier $g_n$.  Unfortunately, we do not know how to measure the entirety of one's brain, much less one's mind. 
% Moreover, even if we did, we do not know the joint distribution of mind/brain pairs, $F_{MB}$.  
We therefore must restrict our interest to a mind/brain \emph{property} pair.  
A mind (mental) property might be a person's intelligence, psychological state, current thought, gender identity, etc.  A brain property might be the number of cells in a person's brain at some time $t$, or the collection of spike trains of all neurons in the brain during some time period $t$ to $t'$.  Regardless of the details of the specifications of the mental property and the brain property, given such specifications, one can assume a model, $\mc{F}$.  We would like a classifier, $g_n$, that is guaranteed to be consistent, no matter which of the possible distributions $F_{MB} \in \mc{F}$ is the true distribution.  A classifier with such a property is called a \emph{universally consistent classifier}.  
Below, under a very general mind-brain model $\mc{F}$, we construct a universally consistent classifier. 
% is readily available.

@@

\begin{thex} \label{exp:1}
Let the physical property under consideration be brain connectivity structure, so $b$ is a brain-graph (``connectome'') with vertices representing neurons (or collections thereof) and edges representing synapses (or collections thereof). Further let $\mB$, the brain observation space, be the collection of all graphs on a given finite number of vertices, and let $\mc{M}$, the mental property observation space, be finite. Now, imagine collecting very large amounts of very accurate identically and independently sampled  brain-graph data and associated mental property indicators from $F_{MB}$. A $k_n$-nearest neighbor classifier using a Frobenius norm is universally consistent (see Methods for details). 
% Therefore, %Theorem \ref{thm:1} applies and 
The existence of a universally consistent classifier guarantees that eventually (in $n,n'$) we will be able to conclude $\MeB$ for this mind-brain property pair, if indeed $\varepsilon$-supervenience holds. This logic holds for directed graphs or multigraphs or hypergraphs with discrete edge weights and vertex attributes, as well as unlabeled graphs (see \cite{VP11_unlabeled} for details). Furthermore, the proof holds for other matrix norms (which might speed up convergence and hence reduce the required $n$), and the regression scenario where $|\mM|$ is infinite (again, see Methods for details).  
\end{thex}
Thus, under the conditions stated in the above \emph{Gedankenexperiment}, we have:
\begin{thm} \label{thm:3}
	$\MeB \implies \beta \conv 1$ as $n,n'\conv \infty$, but there is no universal $n,n'$ that guarantes $\beta=1$.
\end{thm}

@@

Theorem \ref{thm:3} demonstrates the availability of a consistent test under certain restrictions.  The arbitrary slow convergence theorem, however, demonstrates that convergence rates might be unbearably slow.  We therefore provide an illustrative example of the feasibility of such a test on synthetic data.

@@

Let $A_{uv}$ be the number of chemical synapses between neuron $u$ and neuron $v$ according to \cite{VarshneyChklovskii09}.  Then, let $\mc{S}$ be the set of edges deemed responsible for odor-evoked behavior according to \cite{ChalasaniBargmann07}.  If odor-evoked behavior is supervenient on this signal subgraph $\mc{S}$, then the distribution of edges in $\mc{S}$ must differ between the two classes of odor evoked behavior \cite{VP11_sigsub}.  Let $E_{uv|j}$ denote the expected number of edges from vertex $v$ to vertex $u$ in class $j$.   For class $m_0$, let $E_{uv|1}=A_{uv}+\eta$,  where $\eta=0.05$ is a small noise parameter  (it is believed that the C.~elegans connectome is similar across organisms \cite{Durbin87}). For class $m_1$, let $E_{uv|1}=A_{uv}+z_{uv}$, where the signal parameter $z_{uv}=\eta$ for all edges not in $\mc{S}$, and $z_{uv}$ is uniformly sampled from $[-5,5]$ for all edges within $\mc{S}$. For both classes, let each edge be Poisson distributed, $F_{A_{uv}|M=m_j}$= Poisson$(E_{uv|j})$.
% Let the mental property correspond to the C.~elegans exhibiting (or not exhibiting) a particular behavior (e.g., response to an odor).  Thus, one can specify $F_{M=m_1}$ and $F_{M=m_0}$, the probability of a C.~elegans exhibiting or not a particular behavior.   The class-conditional distributions $F_{B|M=m_j}$ are therefore the distributions of brain-graphs conditional on whether the C.~elegans exhibits said property.

@@

We consider $k_n$-nearest neighbor classification of labeled multigraphs (directed, with loops) on 279 vertices, under Frobenius norm. The $k_n$-nearest neighbor classifier used here satisfies $k_n \rightarrow \infty$ as $n \rightarrow \infty$ and $k_n/n \rightarrow 0$ as $n \rightarrow \infty$, ensuring universal consistency. (Better classifiers can be constructed for the joint distribution $F_{MB}$ used here; however, we demand universal consistency.)  Figure \ref{fig:1} shows that for this simulation, rejecting $(\eps=0.1)$-supervenience at $\alpha=0.01$ only requires a few hundred training samples.

@@

\begin{figure}[!ht]
\centering 
\includegraphics[width=.5\linewidth]{Lhatplot}
\caption{C.~elegans graph classification simulation results.  The estimated hold-out misclassification rate $\hL^{n'}_{F}(g_{n})$  (with $n'=1000$ testing samples) %$\hL^{1000}_{\PP}(g_{\mt{n}})$ 
is plotted as a function of class-conditional training sample size $n_j=n/2$, suggesting that for $\varepsilon=0.1$ we can determine that $\MeB$ holds with $99\%$ confidence with just a few hundred training samples generated from $F_{MB}$. Each dot depicts $\hL^{n'}_{F}(g_n)$ for some $n$; standard errors are $(\hL^{n'}_{F}(g_{n}) (1-\hL^{n'}_{F}(g_{n}))/n')^{1/2}$.  For example, at $n_j = 150$ we have $k_n = \lfloor\sqrt{n}\rfloor=11$ (where $\lfloor\cdot\rfloor$ indicates the floor operator), $\hL^{n'}_{F}(g_{n}) = 0.069$, and standard error less than $0.01$. We reject $H_0^{0.1}: L^* \geq 0.1$ at $\alpha=0.01$. Note that $L^* \approx 0$ for this simulation.}
\label{fig:1}
\end{figure}

@@

Alas, arbitrary slow convergence theorems demonstrate that there is no universal $n,n'$ for which convergence is guaranteed.  Thus, a failure to reject is ambiguous: even if the data satisfy the above assumptions, the failure to reject may be due to either (i) an insufficient amount of data or (ii) $\mc{M}$ may not be $\eps$-supervenient on $\mc{B}$.  Moreover, the data will not, in general, satisfy the above assumptions.  In addition to dependence (because each human does not exist in a vacuum), the mental property measurements will often be ``noisy'' (for example, accurately diagnosing psychiatric disorders is a sticky wicket \cite{Kessler2005}). 
% has a number of possible explanations.  First, the amount of data might be insufficient; collecting more data or utilizing a more informative distance might resolve this difficulty.  Second, the class labels (mental property indicators) may not be faithful representations of the mental property of interest (mental property observations may be ``noisy'' indicators). 
% 
% Second, it may be that the mental property under investigation is supervenient to a different brain property.  
% Third,  Fourth, the mental property may really not be supervenient to any brain properties (although testing this more global notion of supervenience is not possible, as one can never be sure that one as measured all brain properties).  
Nonetheless, synthetic data analysis suggests that under somewhat realistic assumptions, convergence obtains with an amount of data one might conceivably collect (Figure \ref{fig:1} and ensuing discussion).

>>>>

\noindent Questions and assumptions about mind-brain supervenience go back at least as far as Plato's dialogues in circa 400 BCE \cite{Plato97}.  While there are many different notions of supervenience, we find Davidson's canonical description particularly illustrative \cite{Davidson70}:
%  The mind-brain supervenience notion was canonized in 1970 with the following quote from Donald Davidson \cite{Davidson70}:
\begin{quotation}
\noindent [mind-brain] supervenience might be taken to mean that there cannot be two events alike in all physical respects but differing in some mental respect, or that an object cannot alter in some mental respect without altering in some physical respect.
\end{quotation}
This philosophical conjecture has potentially widespread implications.  
% We consider a special case of supervenience of considerable interest.  Specifically, this work addresses a version of a local supervenience between mental properties and brain properties, with particular emphasize on brain-graphs (i.e., connectivity structure).  Supervenience questions or assumptions are fundamental to 
% The determination of supervenience (or lack thereof) of minds on brains has potentially important implications in 
% a number of fields of inquiry.  
For example, neural network theory and artificial intelligence often implicitly assume 
% take a generalized notion of 
a local version mind-brain supervenience 
% as an assumption 
% , which if falsified, might change modern approaches to learning 
\cite{Haykin2008,Ripley2008}. Cognitive neuroscience similarly seems to operate under such assumptions
% , which if falsified, might result in novel perspectives and theories 
\cite{Gazzaniga2008}.  Philosophers continue to debate and refine notions of supervenience 
% And the question of mind-brain supervenience continues to be debated amongst philosophers 
\cite{Kim2007}.  
% Moreover, many cognitive neuroscience investigations could be cast as questions
Yet, to date, relatively scant attention has been paid to what might be empirically learned about supervenience.

@@

% Ideally, 
Based on the above considerations,
we might consider the following hypothesis test: $H_0: L^*>0$ and $H_A: L^*=0$; rejecting the null indicates that $\MsB$. Unfortunately, 
% because the null hypothesis is not closed, 
the alternative hypothesis lies on the boundary, so the $p$-value is always equal to unity \cite{Bickel2000}.  From this, Theorem \ref{thm:2} follows immediately:
\begin{thm} \label{thm:2}
	There does not exist a viable test of $\MsB$.
\end{thm}

@@

\subsection*{Conditions for a consistent statistical test for $\eps$-supervenience} % (fold)

@@

\subsection*{The existence and construction of a consistent statistical test for $\eps$-supervenience} % (fold)
\label{sub:uc}

The above considerations indicate the existence of a consistent test for $\eps$-supervenience whenever the classifier used is consistent.  
% that to construct a statistical test with guaranteed limiting unity power, a universally consistent classifier (one for which $\hL^{n'}_F(g_n) \conv L_F(g^*)$ as $n,n' \conv \infty$) is required. 
To actually implement such a test, one must be able to (i) measure mind/brain pairs and (ii) have a consistent classifier $g_n$.  Unfortunately, we do not know how to measure the entirety of one's brain, much less one's mind. 
% Moreover, even if we did, we do not know the joint distribution of mind/brain pairs, $F_{MB}$.  
We therefore must restrict our interest to a mind/brain \emph{property} pair.  
A mind (mental) property might be a person's intelligence, psychological state, current thought, gender identity, etc.  A brain property might be the number of cells in a person's brain at some time $t$, or the collection of spike trains of all neurons in the brain during some time period $t$ to $t'$.  Regardless of the details of the specifications of the mental property and the brain property, given such specifications, one can assume a model, $\mc{F}$.  We desire a classifier $g_n$ that is guaranteed to be consistent, no matter which of the possible distributions $F_{MB} \in \mc{F}$ is the true distribution.  A classifier with such a property is called a \emph{universally consistent classifier}.  
Below, under a very general mind-brain model $\mc{F}$, we construct a universally consistent classifier. 
% is readily available.

@@

\begin{thex} \label{exp:1}
Let the physical property under consideration be brain connectivity structure, so $b$ is a brain-graph (``connectome'') with vertices representing neurons (or collections thereof) and edges representing synapses (or collections thereof). Further let $\mB$, the brain observation space, be the collection of all graphs on a given finite number of vertices, and let $\mc{M}$, the mental property observation space, be finite. Now, imagine collecting very large amounts of very accurate identically and independently sampled  brain-graph data and associated mental property indicators from $F_{MB}$. A $k_n$-nearest neighbor classifier using a Frobenius norm is universally consistent (see Methods for details). 
% Therefore, %Theorem \ref{thm:1} applies and 
The existence of a universally consistent classifier guarantees that eventually (in $n,n'$) we will be able to conclude $\MeB$ for this mind-brain property pair, if indeed $\varepsilon$-supervenience holds. This logic holds for directed graphs or multigraphs or hypergraphs with discrete edge weights and vertex attributes, as well as unlabeled graphs (see \cite{VP11_unlabeled} for details). Furthermore, the proof holds for other matrix norms (which might speed up convergence and hence reduce the required $n$), and the regression scenario where $|\mM|$ is infinite (again, see Methods for details).  
\end{thex}
Thus, under the conditions stated in the above \emph{Gedankenexperiment}, universal consistency yields:
\begin{thm} \label{thm:3}
	$\MeB \implies \beta \conv 1$ as $n,n'\conv \infty$.
\end{thm}

Unfortunately, the rate of convergence of $L_{F}(g_n)$ to $L_{F}(g^*)$ depends on the (unknown) distribution $F=F_{MB}$ \cite{DGL96}. Furthermore, arbitrarily slow convergence theorems regarding the rate of convergence of $L_{F}(g_n)$ to $L_{F}(g^*)$ demonstrate that there is no universal $n,n'$ which will guarantee that the test has power greater than any specified target $\beta > \alpha$ \cite{Devroye83}. For this reason, the test outlined above can provide only a one-sided conclusion: if we reject we can be $100(1-\alpha)$\% confident that $\MeB$ holds, but we can never be confident in its negation; rather, it may be the case that the evidence in favor of $\MeB$ is insufficient 
% for any number of reasons, including that 
because we simply have not yet collected enough data. 
% Thus, without restrictions on $F_{MB}$, arbitrarily slow convergence theorems imply that our theorem of $\varepsilon$-supervenience does not \tr{strictly} satisfy Popper's {\it falsifiability} requirement \cite{Popper}.
This leads immediately to the following theorem:
\begin{thm} \label{thm:4}
There is no universal $n,n'$ that guarantees $\beta \geq \beta_{min}>\alpha$.
\end{thm}

@@

Theorem \ref{thm:3} demonstrates the availability of a consistent test under certain restrictions.  Theorem \ref{thm:4}, however, demonstrates that convergence rates might be unbearably slow.  We therefore provide an illustrative example of the feasibility of such a test on synthetic data.

@@

Let $A_{uv}$ be the number of chemical synapses between neuron $u$ and neuron $v$ according to \cite{VarshneyChklovskii09}.  Then, let $\mc{S}$ be the set of edges deemed responsible for odor-evoked behavior according to \cite{ChalasaniBargmann07}.  If odor-evoked behavior is supervenient on this signal subgraph $\mc{S}$, then the distribution of edges in $\mc{S}$ must differ between the two classes of odor evoked behavior \cite{VP11_sigsub}.  Let $E_{uv|j}$ denote the expected number of edges from vertex $v$ to vertex $u$ in class $j$.   For class $m_0$, let $E_{uv|0}=A_{uv}+\eta$,  where $\eta=0.05$ is a small noise parameter  (it is believed that the C.~elegans connectome is similar across organisms \cite{Durbin87}). For class $m_1$, let $E_{uv|1}=A_{uv}+z_{uv}$, where the signal parameter $z_{uv}=\eta$ for all edges not in $\mc{S}$, and $z_{uv}$ is uniformly sampled from $[-5,5]$ for all edges within $\mc{S}$. For both classes, let each edge be Poisson distributed, $F_{A_{uv}|M=m_j}$= Poisson$(E_{uv|j})$.
% Let the mental property correspond to the C.~elegans exhibiting (or not exhibiting) a particular behavior (e.g., response to an odor).  Thus, one can specify $F_{M=m_1}$ and $F_{M=m_0}$, the probability of a C.~elegans exhibiting or not a particular behavior.   The class-conditional distributions $F_{B|M=m_j}$ are therefore the distributions of brain-graphs conditional on whether the C.~elegans exhibits said property.

@@

We consider $k_n$-nearest neighbor classification of labeled multigraphs (directed, with loops) on 279 vertices, under Frobenius norm. The $k_n$-nearest neighbor classifier used here satisfies $k_n \rightarrow \infty$ as $n \rightarrow \infty$ and $k_n/n \rightarrow 0$ as $n \rightarrow \infty$, ensuring universal consistency. (Better classifiers can be constructed for the joint distribution $F_{MB}$ used here; however, we demand universal consistency.)  Figure \ref{fig:1} shows that for this simulation, rejecting $(\eps=0.1)$-supervenience at $\alpha=0.01$ requires only a few hundred training samples.

@@

\begin{figure}[!ht]
\centering 
\includegraphics[width=.5\linewidth]{Lhatplot}
\caption{C.~elegans graph classification simulation results.  The estimated hold-out misclassification rate $\hL^{n'}_{F}(g_{n})$  (with $n'=1000$ testing samples) %$\hL^{1000}_{\PP}(g_{\mt{n}})$ 
is plotted as a function of class-conditional training sample size $n_j=n/2$, suggesting that for $\varepsilon=0.1$ we can determine that $\MeB$ holds with $99\%$ confidence with just a few hundred training samples generated from $F_{MB}$. Each dot depicts $\hL^{n'}_{F}(g_n)$ for some $n$; standard errors are $(\hL^{n'}_{F}(g_{n}) (1-\hL^{n'}_{F}(g_{n}))/n')^{1/2}$.  For example, at $n_j = 180$ we have $k_n = \lfloor\sqrt{8 n}\rfloor=53$ (where $\lfloor\cdot\rfloor$ indicates the floor operator), $\hL^{n'}_{F}(g_{n}) = 0.057$, and standard error less than $0.01$. We reject $H_0^{0.1}: L^* \geq 0.1$ at $\alpha=0.01$. Note that $L^* \approx 0$ for this simulation.}
\label{fig:1}
\end{figure}

@@

Alas, arbitrary slow convergence theorems demonstrate that there is no universal $n,n'$ for which convergence is guaranteed (Theorem \ref{thm:4}).  Thus, a failure to reject is ambiguous: even if the data satisfy the above assumptions, the failure to reject may be due to either (i) an insufficient amount of data or (ii) $\mc{M}$ may not be $\eps$-supervenient on $\mc{B}$.  Moreover, the data will not, in general, satisfy the above assumptions.  In addition to dependence (because each human does not exist in a vacuum), the mental property measurements will often be ``noisy'' (for example, accurately diagnosing psychiatric disorders is a sticky wicket \cite{Kessler2005}). 
% has a number of possible explanations.  First, the amount of data might be insufficient; collecting more data or utilizing a more informative distance might resolve this difficulty.  Second, the class labels (mental property indicators) may not be faithful representations of the mental property of interest (mental property observations may be ``noisy'' indicators). 
% 
% Second, it may be that the mental property under investigation is supervenient to a different brain property.  
% Third,  Fourth, the mental property may really not be supervenient to any brain properties (although testing this more global notion of supervenience is not possible, as one can never be sure that one as measured all brain properties).  
Nonetheless, synthetic data analysis suggests that under somewhat realistic assumptions, convergence obtains with an amount of data one might conceivably collect (Figure \ref{fig:1} and ensuing discussion).
